---
output: rmarkdown::github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file    -->

## Demo for `tools::md5sum()`

The `tools::md5sum()` function computes the 32-byte MD5 hashes of one or more files. 

From the `md5sum` help page:

> A MD5 ‘hash’ or ‘checksum’ or ‘message digest’ is a 128-bit summary of the file contents represented by 32 hexadecimal digits. Files with different MD5 sums are different: only very exceptionally (and usually with the intent to deceive) are those with the same sums different.

> On Windows all files are read in binary mode (as the md5sum utilities there do): on other OSes the files are read in the default mode (almost always text mode where there is more than one).

> MD5 sums are used as a check that R packages have been unpacked correctly and not subsequently accidentally modified.


### Step 0: Load packages

The `md5sum()` function is in the `tools` package. 

```{r setup}
library(tools) 
library(data.table)
```


### Step 1: Create some source files

Write some source test files to a temporary directory. These are the files for which we will create the source (reference) md5 hashes.
```{r writeFiles}
# Create a temporary directory for the source files. 
tmpDir <- paste0(tempdir(), "/testArea")

# Create a directory structure with subfolders
dir.create(paste(tmpDir, "subf1", "subf2", sep = "/"), recursive = TRUE)

# Write some text and binary files
write.csv(PlantGrowth, file = paste(tmpDir, "PlantGrowth.csv", sep = "/"))
write.csv(ChickWeight, file = paste(tmpDir, "subf1", "ChickWeight.csv", sep = "/"))
write.csv(trees, file = paste(tmpDir, "subf1", "subf2", "trees.csv", sep = "/"))
write.csv(volcano, file = paste(tmpDir, "subf1", "subf2", "volcano.csv", sep = "/"))
saveRDS(ChickWeight, file = paste(tmpDir, "subf1", "ChickWeight.Rdata", sep = "/"))
saveRDS(trees, file = paste(tmpDir, "subf1", "subf2", "trees.Rdata", sep = "/"))
```

List the source files.

```{r listFiles}
listTmpFiles <- list.files(tmpDir, full.names = FALSE, recursive = TRUE)
listTmpFiles
```

### Step 2: Create source MD5 hashes dataset

Create a dataset with MD5 check sums for the source files.
```{r md5sumDataset, R.options = list(width = 120)}
wdOrig <- getwd()
setwd(tmpDir)
MD5.orig <- data.table(fileName = listTmpFiles)[
  , md5Orig := lapply(fileName, md5sum)][
  , md5OrigText := paste0(md5Orig, " ", fileName)][
    , .(md5OrigText)]
MD5.orig
```

Write source MD5 hashes followed by the respective file name to a `.md5` file.

```{r writeMd5File}
md5Filename <- paste0(tmpDir, "/", "orig.md5")
writeLines(text = MD5.orig$md5OrigText, con = md5Filename)
```

> Note: On *nix systems the file `orig.md5` can be check using `md5sum -c orig.md5`.

### Step 3: Create a corrupted version of source files to check

To simulate a user receiving a corrupted version of the source files, let's intentionally make some changes to the source files to get MD5 hash errors during the checks below.

Change a file name to get an error when we check the MD5 hashes below.
```{r changeFiles1}
file.rename(from = paste(tmpDir, "subf1", "subf2", "trees.csv", sep = "/"),
            to = paste(tmpDir, "subf1", "subf2", "TREES.csv", sep = "/"))
```
Change the content of another file to get an error when we check the MD5 hashes below.
```{r changeFiles2}
file.append(paste(tmpDir, "subf1", "ChickWeight.csv", sep = "/"), 
            paste(tmpDir, "PlantGrowth.csv", sep = "/"))
setwd(wdOrig)
```

These are the corrupted files that the user will actually receive.
```{r listCorruptedFiles}
listTmpCorruptedFiles <- list.files(tmpDir, full.names = FALSE, recursive = TRUE)
listTmpCorruptedFiles
```

### Step 4: Generate the MD5 hashes for the corrupted files

```{r md5sumCheck}
setwd(tmpDir)
# Check all files except the orig.md5 file. 
checkFiles <- list.files(tmpDir, full.names = FALSE, recursive = TRUE, 
                         pattern = "[^orig.md5]")
checkFiles

MD5.check <- data.table(fileName = checkFiles)
MD5.check[, md5Check := lapply(fileName, md5sum)]
MD5.check
```

### Step 5: Read in source MD5 check sums text file

Reading in the MD5 hashes for the source files from `orig.md5` that we created above. 
```{r md5sumOrig}
origFileMD5 <- fread(md5Filename, header = FALSE, col.names = c("md5Orig", "fileName"))
origFileMD5
```

### Step 6: Check the MD5 hashes

Verify the hashes from the `orig.md5` with the hashes from the "check" MD5 sums. 
```{r md5sumVerify, R.options = list(width = 100)}
MD5verify <- merge(origFileMD5, MD5.check, by = "fileName", all = TRUE)
MD5verify[, Result := fifelse(md5Orig == md5Check, "OK", "Error")]

# Print out the results
MD5verify
```

+ `OK`: Source and received files match.
+ `Error`: Corrupted file or file not found.
+ `NA`: File received but it is not in the source files. 

---

The R code for this README file is available at:  [Demo_for_tools_md5sum.R](https://github.com/saghirb/md5sum_R_Demo/Demo_for_tools_md5sum.R)

